---
title: '02 Cell Typing - `r unlist(strsplit(getwd(), "/"))[6]`'
author:
  - name: "Emir Turkes [emir.turkes@eturkes.com]"
  - name: "UK Dementia Research Institute at UCL"
date: '`r strftime(Sys.time(), "%B %d, %Y")`'
link-citations: true
output:
  html_document:
    code_folding: hide
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(
    "..", "..", "results", unlist(strsplit(getwd(), "/"))[6], "02_cell_typing.html"
  ))})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
h3 {font-size: 20px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {font-size: 0.95em; padding-left: 25px; text-indent: 0;}
.tocify-subheader .tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 35px; text-indent: 0;
}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of [FAN1 KO iPSC snRNAseq](https://github.com/eturkes/FAN1KO-ipsc-snRNAseq).*

The data here will be referenced using the name ``r unlist(strsplit(getwd(), "/"))[6]``.

```{r}
#    This file is part of FAN1KO-ipsc-snRNAseq.
#    Copyright (C) 2022  Emir Turkes, Emma Bunting, Jessica Olive,
#    Jasmine Donaldson, Sarah Tabrizi, UK Dementia Research Institute at
#    UCL
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

packages <- c(
  "conflicted", "Seurat", "DT", "dplyr", "ggrepel", "ggplot2", "GSVA", "patchwork", "SingleR", "BiocParallel",
  "parallel"
)
invisible(suppressPackageStartupMessages(lapply(packages, library, character.only = TRUE)))
source(file.path("..", "utils.R"))

`%notin%` <- Negate(`%in%`)

analysis_no <- 2
data_name <- unlist(strsplit(getwd(), "/"))[6] # Name of dataset.
data_dir <- file.path("..", "..", "data") # Backed up data.
results_dir <- file.path("..", "..", "results")

# Unique cache directory for each analysis number.
# ------------------------------------------------
cache_dir <- file.path("..", "..", "cache", data_name, paste0("0", analysis_no))
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}
# ------------------------------------------------

knitr::opts_chunk$set(fig.width = 10, fig.height = 7, dpi = 150)
```

# Prep

```{r}
seurat <- readRDS(file.path(cache_dir, "..", "01", "all_seurat.rds"))
seurat

subset_list <- list(seurat)
subset_names <- "all"

red_dim_plot(seurat, "umap1", "umap2", "seurat_clusters", "cat")
```

# Genes of Interest

```{r}
genes <- read.delim(file.path(data_dir, "gene-sets", "genes-of-interest.tsv"))
datatable_download(genes)
```

## MMR

```{r, fig.width = 15, fig.height = 14}
# gene_plot(subset_list, subset_names, trimws(genes$MMR[nzchar(genes$MMR)]))

gene_plot(subset_list, subset_names, c(trimws(genes$MMR[nzchar(genes$MMR)]), "HTT", "FAN1", "EXO1", trimws(genes$Housekeeping[nzchar(genes$Housekeeping)])[-3]))
```

## Other

```{r, fig.height = 6, fig.width = 10}
gene_plot(subset_list, subset_names, trimws(genes$Other[nzchar(genes$Other)]))
```

## Neuronal

```{r, fig.height = 6, fig.width = 10}
gene_plot(subset_list, subset_names, trimws(genes$Neuronal[nzchar(genes$Neuronal)]))
```

## MSN

```{r, fig.height = 6, fig.width = 15}
gene_plot(subset_list, subset_names, trimws(genes$MSN[nzchar(genes$MSN)]))
```

## Astrocytic

```{r, fig.height = 6}
gene_plot(subset_list, subset_names, trimws(genes$Astrocytic[nzchar(genes$Astrocytic)]))
```

## Proliferative

```{r, fig.height = 6, fig.width = 10}
gene_plot(subset_list, subset_names, trimws(genes$Proliferative[nzchar(genes$Proliferative)]))
```

## Housekeeping

```{r, fig.height = 8}
gene_plot(subset_list, subset_names, trimws(genes$Housekeeping[nzchar(genes$Housekeeping)]))
```

# Individual Marker Genes

## Smith-Geater

```{r}
smith_markers <- vector("list", length = 7)
names(smith_markers) <- c(
  "mature neuron", "D1 and D2+ spiny neurons", "NSC", "D1+ spiny and cholinergic striatal neurons",
  "D1+ spiny neurons", "NPCs", "immature neurons"
)
smith_markers[[1]] <- c("TPD52L1", "MEF2C", "PAX6", "SIX3")
smith_markers[[2]] <- c("MEIS2", "DLX5", "HPCA")
smith_markers[[3]] <- c("NES", "CCND1")
smith_markers[[4]] <- c("DLX5", "ISL1", "RAC3", "ARG2")
smith_markers[[5]] <- c("MEIS2", "PPP1R1A", "CAMKV", "RAC3", "GRIN2B")
smith_markers[[6]] <- c("CCND1", "HOMER3", "EBF1")
smith_markers[[7]] <- c("VGF", "H1F0")

remove <- unique(unlist(smith_markers))[unique(unlist(smith_markers)) %notin% VariableFeatures(seurat)]
smith_markers_highvar <- lapply(smith_markers, function(x) x[!(x %in% remove)])

for (i in seq_along(smith_markers_highvar)) {
  if (any(smith_markers_highvar[[i]] %in% rownames(seurat))) {
    print(
      FeaturePlot(seurat, smith_markers_highvar[[i]], order = TRUE, cols = c("lightgrey", "red")) +
        plot_annotation(
          names(smith_markers_highvar)[i],
          theme = theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"))
        )
    )
  }
}
```

## Tran

```{r}
tran_markers <- read.delim(file.path(data_dir, "gene-sets", "tran_markers.tsv"))
tran_markers <- as.list(tran_markers)

remove <- unique(unlist(tran_markers))[unique(unlist(tran_markers)) %notin% VariableFeatures(seurat)]
tran_markers_highvar <- lapply(tran_markers, function(x) x[!(x %in% remove)])

tran_markers_expanded <- read.delim(file.path(data_dir, "gene-sets", "tran_markers_expanded.tsv"))
tran_markers_expanded <- as.list(tran_markers_expanded)

tran_markers_aggregated <- tran_markers_expanded
names(tran_markers_aggregated) <- gsub("(.+?)(\\_.*)", "\\1", names(tran_markers_expanded))
tran_markers_aggregated <- tapply(
  unlist(tran_markers_aggregated, use.names = FALSE),
  rep(names(tran_markers_aggregated), lengths(tran_markers_aggregated)),
  c
)
tran_markers_aggregated <- lapply(tran_markers_aggregated, function(x) x[!(x == "")])

remove <-
  unique(unlist(tran_markers_aggregated))[unique(unlist(tran_markers_aggregated)) %notin% VariableFeatures(seurat)]
tran_markers_aggregated_highvar <- lapply(tran_markers_aggregated, function(x) x[!(x %in% remove)])

for (i in seq_along(tran_markers_aggregated_highvar)) {
  if (any(tran_markers_aggregated_highvar[[i]] %in% rownames(seurat))) {
    print( # Only plot first four genes.
      FeaturePlot(seurat, tran_markers_aggregated_highvar[[i]][1:4], order = TRUE, cols = c("lightgrey", "red")) +
        plot_annotation(
          names(tran_markers_aggregated_highvar)[i],
          theme = theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"))
        )
    )
  }
}
```

## Tran Figure S4 Markers

```{r}
tran_markers_figS4 <- read.delim(file.path(data_dir, "gene-sets", "tran_markers_figS4.tsv"), header = FALSE)
tran_markers_figS4 <- tran_markers_figS4[tran_markers_figS4$V1 %in% VariableFeatures(seurat), ]

for (i in seq_along(tran_markers_figS4)) {
  print(FeaturePlot(seurat, tran_markers_figS4[i], order = TRUE, cols = c("lightgrey", "red")))
}
```

## Bhaduri

```{r}
bhaduri_markers <- read.delim(file.path(data_dir, "gene-sets", "curated-development-genes.tsv"), check.names = FALSE)

bhaduri_markers <- bhaduri_markers[ , 1:which(colnames(bhaduri_markers) == "Polioudakis Neuron 2019")]
bhaduri_markers <- bhaduri_markers[ , -(which(colnames(bhaduri_markers) == "oRG vRG shared genes"))]
bhaduri_markers <- Filter(function(x) !all(is.na(x)), bhaduri_markers)
bhaduri_markers <- as.list(bhaduri_markers)
bhaduri_markers <- lapply(bhaduri_markers, function(x) x[!(x == "")])

remove <- c(
  "SORCS1", "ADRA2A", "SLA", "STMN2", "SORBS2", "ELAVL2", "DCX", "ZFHX4", "GLI3", "CD99", "IGFBP7",
  "AURKC", "HMGB1", "HMGB3", "CTNNB1", "HES1", "VIM", "PTN", "CLU", "IL33", "NPR3", "CCNB2",
  "KIF15", "ARHGAP11B", "CENPK", "NEUROD2", "NEUROD6", "DLX2", "ERBB4", "SOX5", "BCL11B"
)
bhaduri_markers <- bhaduri_markers
for (i in seq_along(remove)) {
  bhaduri_markers <- rapply(bhaduri_markers, function(x) ifelse(x == remove[i], "", x), how = "replace")
}
bhaduri_markers <- lapply(bhaduri_markers, function(x) x[!(x == "")])
remove <- which(
  names(bhaduri_markers) %in% c(
    "Radial Glia_vRG", "Radial Glia_tRG", "Microglia", "oRG specific genes", "vRG specific genes"
  )
)
bhaduri_markers <- bhaduri_markers[-remove]

remove <- unique(unlist(bhaduri_markers))[unique(unlist(bhaduri_markers)) %notin% VariableFeatures(seurat)]
bhaduri_markers_highvar <- lapply(bhaduri_markers, function(x) x[!(x %in% remove)])

for (i in seq_along(bhaduri_markers)) {
  if (any(bhaduri_markers[[i]] %in% rownames(seurat))) {
    print(
      FeaturePlot(seurat, bhaduri_markers[[i]], order = TRUE, cols = c("lightgrey", "red")) +
        plot_annotation(
          names(bhaduri_markers)[i],
          theme = theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"))
        )
    )
  }
}
```

# Gene Sets

## Smith-Geater

```{r}
rds <- file.path(cache_dir, "smith_gsva_highvar.rds")
if (file.exists(rds)) {
  gsva <- readRDS(rds)
} else {
  gsva <- gsva(
    as.matrix(GetAssayData(seurat)), smith_markers_highvar, method = "ssgsea",
    ssgsea.norm = FALSE, verbose = FALSE, BPPARAM = MulticoreParam(detectCores())
  )
  saveRDS(gsva, rds)
}
gsva <- (2 * (gsva - min(gsva)) / (max(gsva) - min(gsva))) - 1

gsva_seurat <- CreateAssayObject(data = gsva)
seurat[["GSVA"]] <- gsva_seurat
DefaultAssay(seurat) <- "GSVA"

for (i in seq(nrow(seurat))) {
  print(
    FeaturePlot(seurat, rownames(seurat)[i], order = TRUE) +
      scale_color_gradientn(colors = c("blue", "lightgray", "red"),  limits = c(-1, 1))
  )
}

DefaultAssay(seurat) <- "SCT"
```

## Tran

```{r}
rds <- file.path(cache_dir, "tran_gsva_aggregated.rds")
if (file.exists(rds)) {
  gsva <- readRDS(rds)
} else {
  gsva <- gsva(
    as.matrix(GetAssayData(seurat)), tran_markers_aggregated, method = "ssgsea",
    ssgsea.norm = FALSE, verbose = FALSE, BPPARAM = MulticoreParam(detectCores())
  )
  saveRDS(gsva, rds)
}
gsva <- (2 * (gsva - min(gsva)) / (max(gsva) - min(gsva))) - 1

gsva_seurat <- CreateAssayObject(data = gsva)
seurat[["GSVA"]] <- gsva_seurat
DefaultAssay(seurat) <- "GSVA"

for (i in seq(nrow(seurat))) {
  print(
    FeaturePlot(seurat, rownames(seurat)[i], order = TRUE) +
      scale_color_gradientn(colors = c("blue", "lightgray", "red"),  limits = c(-1, 1))
  )
}

DefaultAssay(seurat) <- "SCT"
```

# SingleR

## Tran

```{r}
ref <- vector("list", length = 5)

load(file.path(data_dir, "tran", "SCE_AMY-n5_tran-etal.rda"))
ref[[1]] <- sce.amy.tran
rm(sce.amy.tran)
unique(ref[[1]]$cellType)
remove <- grep("drop", ref[[1]]$cellType)
ref[[1]] <- ref[[1]][ , -remove]
ref[[1]]$cellType <- factor(ref[[1]]$cellType)
ref[[1]]$broad_cellType <- factor(sub("_.*", "", ref[[1]]$cellType))
unique(ref[[1]]$cellType)
unique(ref[[1]]$broad_cellType)

load(file.path(data_dir, "tran", "SCE_DLPFC-n3_tran-etal.rda"))
ref[[2]] <- sce.dlpfc.tran
rm(sce.dlpfc.tran)
ref[[2]]$cellType <- factor(ref[[2]]$cellType)
ref[[2]]$broad_cellType <- factor(sub("_.*", "", ref[[2]]$cellType))
unique(ref[[2]]$cellType)
unique(ref[[2]]$broad_cellType)

load(file.path(data_dir, "tran", "SCE_HPC-n3_tran-etal.rda"))
ref[[3]] <- sce.hpc.tran
rm(sce.hpc.tran)
unique(ref[[3]]$cellType)
remove <- grep("drop", ref[[3]]$cellType)
ref[[3]] <- ref[[3]][ , -remove]
ref[[3]]$cellType <- factor(ref[[3]]$cellType)
ref[[3]]$broad_cellType <- factor(sub("_.*", "", ref[[3]]$cellType))
unique(ref[[3]]$cellType)
unique(ref[[3]]$broad_cellType)

load(file.path(data_dir, "tran", "SCE_NAc-n8_tran-etal.rda"))
ref[[4]] <- sce.nac.tran
rm(sce.nac.tran)
unique(ref[[4]]$cellType)
remove <- grep("drop", ref[[4]]$cellType)
ref[[4]] <- ref[[4]][ , -remove]
ref[[4]]$cellType <- factor(ref[[4]]$cellType)
ref[[4]]$broad_cellType <- factor(sub("_.*", "", ref[[4]]$cellType))
unique(ref[[4]]$cellType)
unique(ref[[4]]$broad_cellType)

load(file.path(data_dir, "tran", "SCE_sACC-n5_tran-etal.rda"))
ref[[5]] <- sce.sacc.tran
rm(sce.sacc.tran)
unique(ref[[5]]$cellType)
remove <- grep("drop", ref[[5]]$cellType)
ref[[5]] <- ref[[5]][ , -remove]
remove <- grep("FAT2", ref[[5]]$cellType)
ref[[5]] <- ref[[5]][ , -remove]
ref[[5]]$cellType <- factor(ref[[5]]$cellType)
ref[[5]]$broad_cellType <- factor(sub("_.*", "", ref[[5]]$cellType))
unique(ref[[5]]$cellType)
unique(ref[[5]]$broad_cellType)
```

### All Cell-types

```{r}
rds <- file.path(cache_dir, "tran_singleR_markers.rds")
if (file.exists(rds)) {
  singleR_markers <- readRDS(rds)
} else {
  singleR_markers <- getClassicMarkers(
    ref, list(ref[[1]]$cellType, ref[[2]]$cellType, ref[[3]]$cellType, ref[[4]]$cellType, ref[[5]]$cellType),
    BPPARAM = MulticoreParam(detectCores())
  )
  saveRDS(singleR_markers, rds)
}

rds <- file.path(cache_dir, "tran_singleR.rds")
if (file.exists(rds)) {
  predictions <- readRDS(rds)
} else {
  predictions <- SingleR(
    GetAssayData(seurat, assay = "RNA"),
    ref,
    list(ref[[1]]$cellType, ref[[2]]$cellType, ref[[3]]$cellType, ref[[4]]$cellType, ref[[5]]$cellType),
    genes = list(singleR_markers, singleR_markers, singleR_markers, singleR_markers, singleR_markers),
    BPPARAM = MulticoreParam(detectCores())
  )
  saveRDS(predictions, rds)
}

seurat$tran_singleR <- factor(predictions$labels)
table(seurat$tran_singleR)
red_dim_plot(seurat, "umap1", "umap2", "tran_singleR", "cat")

seurat$tran_singleR_broad <- factor(sub("_.*", "", seurat$tran_singleR))
table(seurat$tran_singleR_broad)
red_dim_plot(seurat, "umap1", "umap2", "tran_singleR_broad", "cat")

remove <- which(
  seurat$tran_singleR_broad == "MSN.D1" | seurat$tran_singleR_broad == "MSN.D2" | seurat$tran_singleR_broad == "Oligo"
)
seurat <- seurat[ , -remove]
table(seurat$tran_singleR_broad)
red_dim_plot(seurat, "umap1", "umap2", "tran_singleR_broad", "cat")
```

### Broad Cell-types

```{r}
rds <- file.path(cache_dir, "tran_singleR_markers_broad.rds")
if (file.exists(rds)) {
  singleR_markers <- readRDS(rds)
} else {
  singleR_markers <- getClassicMarkers(
    ref,
    list(
      ref[[1]]$broad_cellType, ref[[2]]$broad_cellType, ref[[3]]$broad_cellType,
      ref[[4]]$broad_cellType, ref[[5]]$broad_cellType
    ),
    BPPARAM = MulticoreParam(detectCores())
  )
  saveRDS(singleR_markers, rds)
}

rds <- file.path(cache_dir, "tran_singleR_broad.rds")
if (file.exists(rds)) {
  predictions <- readRDS(rds)
} else {
  predictions <- SingleR(
    GetAssayData(seurat, assay = "RNA"),
    ref,
    list(
      ref[[1]]$broad_cellType, ref[[2]]$broad_cellType, ref[[3]]$broad_cellType,
      ref[[4]]$broad_cellType, ref[[5]]$broad_cellType
    ),
    genes = list(singleR_markers, singleR_markers, singleR_markers, singleR_markers, singleR_markers)
  )
  saveRDS(predictions, rds)
}

seurat$tran_singleR <- factor(predictions$labels)
table(seurat$tran_singleR)
red_dim_plot(seurat, "umap1", "umap2", "tran_singleR", "cat")
```

### NAc Only

```{r}
rds <- file.path(cache_dir, "tran_singleR_markers_NAc.rds")
if (file.exists(rds)) {
  singleR_markers <- readRDS(rds)
} else {
  singleR_markers <- getClassicMarkers(
    list(ref[[4]]), list(ref[[4]]$broad_cellType), BPPARAM = MulticoreParam(detectCores())
  )
  saveRDS(singleR_markers, rds)
}

rds <- file.path(cache_dir, "tran_singleR_NAc.rds")
if (file.exists(rds)) {
  predictions <- readRDS(rds)
} else {
  predictions <- SingleR(
    GetAssayData(seurat, assay = "RNA"), list(ref[[4]]), list(ref[[4]]$broad_cellType),
    genes = list(singleR_markers), BPPARAM = MulticoreParam(detectCores())
  )
  saveRDS(predictions, rds)
}

seurat$tran_singleR <- factor(predictions$labels)
table(seurat$tran_singleR)
red_dim_plot(seurat, "umap1", "umap2", "tran_singleR", "cat")
```

# References

This is the concluding section of the document. Here we output the `sessionInfo` and create a bibliography for works cited.

```{r}
sessionInfo()
```
