---
title: '01 DEGs - `r unlist(strsplit(getwd(), "/"))[6]`'
author:
  - name: "Emir Turkes [emir.turkes@eturkes.com]"
  - name: "UK Dementia Research Institute at UCL"
date: '`r strftime(Sys.time(), "%B %d, %Y")`'
link-citations: true
output:
  html_document:
    code_folding: hide
    number_sections: true
    theme: lumen
    highlight: haddock
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(
    "..", "..", "results", unlist(strsplit(getwd(), "/"))[6], "01_DEGs.html"
  ))})
---

<style type="text/css">
body {font-size: 16px;}
h1.title {font-size: 35px;}
h1 {font-size: 24px;}
h2 {font-size: 22px;}
h3 {font-size: 20px;}
.toc-content {padding-left: 0px; padding-right: 0px;}
div.tocify {width: 100%;}
.tocify-subheader .tocify-item {font-size: 0.95em; padding-left: 25px; text-indent: 0;}
.tocify-subheader .tocify-subheader .tocify-item {
  font-size: 0.95em; padding-left: 35px; text-indent: 0;
}
div.main-container {max-width: none; width: 100%;}
</style>

*This file is a part of [FAN1 KO iPSC snRNAseq](https://github.com/eturkes/FAN1KO-ipsc-snRNAseq).*

The data here will be referenced using the name ``r unlist(strsplit(getwd(), "/"))[6]``.

```{r}
#    This file is part of FAN1KO-ipsc-snRNAseq.
#    Copyright (C) 2022  Emir Turkes, Emma Bunting, Jessica Olive,
#    Jasmine Donaldson, Sarah Tabrizi, UK Dementia Research Institute at
#    UCL
#
#    This program is free software: you can redistribute it and/or modify
#    it under the terms of the GNU General Public License as published by
#    the Free Software Foundation, either version 3 of the License, or
#    (at your option) any later version.
#
#    This program is distributed in the hope that it will be useful,
#    but WITHOUT ANY WARRANTY; without even the implied warranty of
#    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
#    GNU General Public License for more details.
#
#    You should have received a copy of the GNU General Public License
#    along with this program.  If not, see <http://www.gnu.org/licenses/>.
#
#    Emir Turkes can be contacted at emir.turkes@eturkes.com

packages <- c("conflicted", "Seurat", "DT", "ggplot2", "clusterProfiler", "org.Hs.eg.db")
invisible(suppressPackageStartupMessages(lapply(packages, library, character.only = TRUE)))
source(file.path("..", "utils.R"))

analysis_no <- 1
data_name <- unlist(strsplit(getwd(), "/"))[6] # Name of dataset.
data_dir <- file.path("..", "..", "data") # Backed up data.
results_dir <- file.path("..", "..", "results")

# Unique cache directory for each analysis number.
# ------------------------------------------------
cache_dir <- file.path("..", "..", "cache", data_name, paste0("0", analysis_no))
if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}
# ------------------------------------------------

knitr::opts_chunk$set(fig.width = 10, fig.height = 7, dpi = 150)
```

# Prep

```{r}
WT <- readRDS(file.path(cache_dir, "..", "..", "WT", "02", "annotated_seurat.rds"))
MUT <- readRDS(file.path(cache_dir, "..", "..", "2H3", "02", "annotated_seurat.rds"))
seurat <- merge(WT, MUT)
rm(WT, MUT)

seurat
```

# MSN DEGs

```{r}
rds <- file.path(cache_dir, "MSN_seurat.rds")
if (file.exists(rds)) {
  sub_seurat <- readRDS(rds)
} else {
  sub_seurat <- subset(seurat, idents = "MSN")
  sub_seurat@active.ident <- factor(sub_seurat$group)
  sub_seurat <- suppressWarnings(SCTransform(sub_seurat, vst.flavor = "v2", verbose = FALSE))
  saveRDS(sub_seurat, rds)
}

rds <- file.path(cache_dir, "MSN_DEGs.rds")
if (file.exists(rds)) {
  DEGs <- readRDS(rds)
} else {
  DEGs <- FindAllMarkers(sub_seurat, verbose = FALSE)
  saveRDS(DEGs, rds)
}

DEGs <- DEGs[DEGs$cluster == "2H3", ]
datatable_download_exp(DEGs)

VlnPlot(sub_seurat, DEGs$gene[seq(12)])
DoHeatmap(sub_seurat, DEGs$gene[seq(50)]) + scale_fill_gradientn(colors = c("blue", "white", "red"))

DEGs_sub <- DEGs[DEGs$avg_log2FC > 0, ]
gse <- enrichGO(DEGs_sub$gene, org.Hs.eg.db, "SYMBOL", "BP", universe = rownames(sub_seurat))
barplot(gse, showCategory = 10) + ggtitle("Upregulated in 2H3")

DEGs_sub <- DEGs[DEGs$avg_log2FC < 0, ]
gse <- enrichGO(DEGs_sub$gene, org.Hs.eg.db, "SYMBOL", "BP", universe = rownames(sub_seurat))
barplot(gse, showCategory = 10) + ggtitle("Downregulated in 2H3")
```

# Inhibitory Neuron DEGs

```{r}
rds <- file.path(cache_dir, "Inhibitory_Neuron_seurat.rds")
if (file.exists(rds)) {
  sub_seurat <- readRDS(rds)
} else {
  sub_seurat <- subset(seurat, idents = "Inhibitory Neuron")
  sub_seurat@active.ident <- factor(sub_seurat$group)
  sub_seurat <- suppressWarnings(SCTransform(sub_seurat, vst.flavor = "v2", verbose = FALSE))
  saveRDS(sub_seurat, rds)
}

rds <- file.path(cache_dir, "Inhibitory_Neuron_DEGs.rds")
if (file.exists(rds)) {
  DEGs <- readRDS(rds)
} else {
  DEGs <- FindAllMarkers(sub_seurat, verbose = FALSE)
  saveRDS(DEGs, rds)
}

DEGs <- DEGs[DEGs$cluster == "2H3", ]
datatable_download_exp(DEGs)

VlnPlot(sub_seurat, DEGs$gene[seq(12)])
DoHeatmap(sub_seurat, DEGs$gene[seq(50)]) + scale_fill_gradientn(colors = c("blue", "white", "red"))

DEGs_sub <- DEGs[DEGs$avg_log2FC > 0, ]
gse <- enrichGO(DEGs_sub$gene, org.Hs.eg.db, "SYMBOL", "BP", universe = rownames(sub_seurat))
barplot(gse, showCategory = 10) + ggtitle("Upregulated in 2H3")

DEGs_sub <- DEGs[DEGs$avg_log2FC < 0, ]
gse <- enrichGO(DEGs_sub$gene, org.Hs.eg.db, "SYMBOL", "BP", universe = rownames(sub_seurat))
barplot(gse, showCategory = 10) + ggtitle("Downregulated in 2H3")
```

# Astrocyte DEGs

```{r}
rds <- file.path(cache_dir, "Astrocyte_seurat.rds")
if (file.exists(rds)) {
  sub_seurat <- readRDS(rds)
} else {
  sub_seurat <- subset(seurat, idents = "Astrocyte")
  sub_seurat@active.ident <- factor(sub_seurat$group)
  sub_seurat <- suppressWarnings(SCTransform(sub_seurat, vst.flavor = "v2", verbose = FALSE))
  saveRDS(sub_seurat, rds)
}

rds <- file.path(cache_dir, "Astrocyte_DEGs.rds")
if (file.exists(rds)) {
  DEGs <- readRDS(rds)
} else {
  DEGs <- FindAllMarkers(sub_seurat, verbose = FALSE)
  saveRDS(DEGs, rds)
}

DEGs <- DEGs[DEGs$cluster == "2H3", ]
datatable_download_exp(DEGs)

VlnPlot(sub_seurat, DEGs$gene[seq(12)])
DoHeatmap(sub_seurat, DEGs$gene[seq(50)]) + scale_fill_gradientn(colors = c("blue", "white", "red"))

DEGs_sub <- DEGs[DEGs$avg_log2FC > 0, ]
gse <- enrichGO(DEGs_sub$gene, org.Hs.eg.db, "SYMBOL", "BP", universe = rownames(sub_seurat))
barplot(gse, showCategory = 10) + ggtitle("Upregulated in 2H3")

DEGs_sub <- DEGs[DEGs$avg_log2FC < 0, ]
gse <- enrichGO(DEGs_sub$gene, org.Hs.eg.db, "SYMBOL", "BP", universe = rownames(sub_seurat))
barplot(gse, showCategory = 10) + ggtitle("Downregulated in 2H3")
```

# OPC DEGs

```{r}
rds <- file.path(cache_dir, "OPC_seurat.rds")
if (file.exists(rds)) {
  sub_seurat <- readRDS(rds)
} else {
  sub_seurat <- subset(seurat, idents = "OPC")
  sub_seurat@active.ident <- factor(sub_seurat$group)
  sub_seurat <- suppressWarnings(SCTransform(sub_seurat, vst.flavor = "v2", verbose = FALSE))
  saveRDS(sub_seurat, rds)
}

rds <- file.path(cache_dir, "OPC_DEGs.rds")
if (file.exists(rds)) {
  DEGs <- readRDS(rds)
} else {
  DEGs <- FindAllMarkers(sub_seurat, verbose = FALSE)
  saveRDS(DEGs, rds)
}

DEGs <- DEGs[DEGs$cluster == "2H3", ]
datatable_download_exp(DEGs)

VlnPlot(sub_seurat, DEGs$gene[seq(12)])
DoHeatmap(sub_seurat, DEGs$gene[seq(50)]) + scale_fill_gradientn(colors = c("blue", "white", "red"))

DEGs_sub <- DEGs[DEGs$avg_log2FC > 0, ]
gse <- enrichGO(DEGs_sub$gene, org.Hs.eg.db, "SYMBOL", "BP", universe = rownames(sub_seurat))
barplot(gse, showCategory = 10) + ggtitle("Upregulated in 2H3")

DEGs_sub <- DEGs[DEGs$avg_log2FC < 0, ]
gse <- enrichGO(DEGs_sub$gene, org.Hs.eg.db, "SYMBOL", "BP", universe = rownames(sub_seurat))
barplot(gse, showCategory = 10) + ggtitle("Downregulated in 2H3")
```

# Mural DEGs

```{r}
rds <- file.path(cache_dir, "Mural_seurat.rds")
if (file.exists(rds)) {
  sub_seurat <- readRDS(rds)
} else {
  sub_seurat <- subset(seurat, idents = "Mural")
  sub_seurat@active.ident <- factor(sub_seurat$group)
  sub_seurat <- suppressWarnings(SCTransform(sub_seurat, vst.flavor = "v2", verbose = FALSE))
  saveRDS(sub_seurat, rds)
}

rds <- file.path(cache_dir, "Mural_DEGs.rds")
if (file.exists(rds)) {
  DEGs <- readRDS(rds)
} else {
  DEGs <- FindAllMarkers(sub_seurat, verbose = FALSE)
  saveRDS(DEGs, rds)
}

DEGs <- DEGs[DEGs$cluster == "2H3", ]
datatable_download_exp(DEGs)

VlnPlot(sub_seurat, DEGs$gene[seq(12)])
DoHeatmap(sub_seurat, DEGs$gene[seq(50)]) + scale_fill_gradientn(colors = c("blue", "white", "red"))

DEGs_sub <- DEGs[DEGs$avg_log2FC > 0, ]
gse <- enrichGO(DEGs_sub$gene, org.Hs.eg.db, "SYMBOL", "BP", universe = rownames(sub_seurat))
barplot(gse, showCategory = 10) + ggtitle("Upregulated in 2H3")

DEGs_sub <- DEGs[DEGs$avg_log2FC < 0, ]
gse <- enrichGO(DEGs_sub$gene, org.Hs.eg.db, "SYMBOL", "BP", universe = rownames(sub_seurat))
barplot(gse, showCategory = 10) + ggtitle("Downregulated in 2H3")
```

# References

This is the concluding section of the document. Here we output the `sessionInfo` and create a bibliography for works cited.

```{r}
sessionInfo()
```
